Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-31 doc/response.tex

\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it equivalent to the definition in Lines 42-43?}

\revtext{Line 46: "......intrinsic generation intervals measure the infectiousness of an
infected individual." This statement is vague. I will get back to this phrase
a few pages later in the discussion on "equilibrium distributions" and
suggest ways to make it more precise.}

We now define intrinsic generation interval properly:

\newtext{The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.}

@@

In Section 2, we also introduce the idea that the intrinsic distributions can be thought of as equilibrium distributions:

\newtext{$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed population at the endemic equilibrium, or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized (forward) generation intervals.}

@@

\revtext{
Lines 44-48: "... the observed generation-interval distribution can change...
...observed generation intervals refer to the time between actual infection
events." The actual infection events correspond to the exact moments of
passing the infection between a pair of infected-susceptible individuals.
Are they observable? In most cases, including contact tracing, observable
events are clinical (e.g. onset of symptoms). The relationship between
the interval defined by the time between actual infection events and the
interval defined by the time between clinical onsets is unclear. At least the
latter depends on the incubation period (of the infectee), in addition to the
latent period and the infectious period (of the infector). The incubation
period is defined as the time from infection to symptom onset, and is
different from the latent period such as in the SEIR model. This paper
does not make such distinction. It assumes that the time between actual
infection events "were observable" in a hypothetical way.
}

\revtext{
Suggestion: Re-phrase "observed generation intervals" or make some further
clarification on this. This is important because naive readers may tend
to take the face value and collect different kinds of observed data and
make inferences, such as that based on clinical events. This is potentially
misleading.
}

We appreciate this point. We now refer to them as \emph{realized} intervals.

\newtext{There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
On the other hand, \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.}

@@

We have also added the following sentence as per suggestions in previous comments:

\newtext{In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
in most cases, observable events are clinical (e.g. onset of symptoms).
There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals \citep{svensson2007note}.
Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use  \citep{britton2019estimation}.
We will not address these complexities here.}

@@

No. See above.

\revtext{5. do you imply that the intrinsic generation interval (not defined in Line 46
but distributed according to Line 101) and the forward measure, and the
objective of the unbiased inference, are all equivalent?}

No. The intrinsic distributions does not account for any spatial effect and therefore are different from the forward distributions.
We have tried to make this clear throughout the manuscript.
In particular, we provide the following sentence in Introduction to distinguish intrinsic from the realized distributions:

\newtext{Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.}

@@

We have shortened and tried to improve this explanation.

\revtext{Lines 241-243: "In a large homogeneously mixing population, the effective
distribution is expected to be equivalent to the intrinsic distribution..."
Define both $g(\tau)$ and gef $g_{eff}(\tau)$ mathematically at the first place and prove
their equivalence in a large homogeneously mixing population.}

We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution):

\newtext{Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.}

@@

\revtext{
One can not even claim the gamma distributions are more realistic
without supported by data. For one thing: all gamma distributions
have exponential tails which may be contested in real data.
}

\revtext{
"... based on more realistic distributions in Appendix." I do
not see "more realistic distributions" more than the Erlang distributions.
How about heavy tailed distributions and distributions with large variances? In fact, many of the discussions in this paper can be generalized
to arbitrarily distributed latent and infectious periods.
}

We had omitted a sentence previous --- we have now clarified this sentence as follows:

\newtext{In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}

@@

The reviewer's discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper. Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.

\revtext{How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests?}

These are hard issues, but beyond the scope of this paper. 
Using exponentials is routine in many applications. 
Using gammas (or Erlangs) generalizes this, and is also commonly done. 
The gamma distribution has been previously tested for many diseases. 
In our case, Figure 5 demonstrates that the gamma approximation of the generation-interval distribution provides a reasonable estimate for the initial reproduction number from the exponential growth rate.
We now discuss additional evidence that gammas often provide reasonable estimates [our approximation framework] as well as potential caveats in assuming a shape of the generation-interval distribution in Discussion.

\newtext{We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously showed that in many cases a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Rini whereas narrower generation-interval distribution leads to higher \Rini) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).}

@@

\revtext{How does the likelihood
function look like? Since the "observed" generation-intervals (assuming
observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
scheme (e.g. retrospective)? Please be explicit.
}

The population-level method does not take into account the observation scheme or the infection process. 
We simply fit the gamma distribution to all realized generation intervals (i.e., the aggregated distribution).
We have clarified the text:

\newtext{The \textbf{population-level method} estimates the initial forward generation-interval distribution by reversing the inverse exponential weighting in the aggregated generation-interval distribution without explicitly accounting for the infection process (i.e., who infected whom) \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau),
\end{equation}
where $r$ is the initial exponential growth rate.
In order to do so, we first approximate the aggregated distribution $a_0$ with a gamma distribution by assuming that realized generation intervals (subject to right-truncation) during the exponential growth phase come from the same gamma distribution;
specifically, we estimate the mean $\bar G$ and shape $\alpha$ of a gamma distribution by maximum likelihood.
Then, the initial forward generation-interval distribution also follows a gamma distribution with mean $\alpha/(\alpha/\bar G - r)$ and shape $\alpha$.
We then use the estimated initial forward generation-interval distribution to infer the initial reproductive number $\Rini$ from the estimated growth rate (using the Euler-Lotka equation).}

@@

We had included this reference in earlier versions of this manuscript a few years ago, thinking that their approach is similar due to their use of Poisson likelihood.
We realized that their approach is quite different from the goal of our paper and have removed the sentence.
They do not use Euler-Lotka or rely on ``observed'' generation intervals;
instead they try to infer a generation-interval distribution from incidence reports.

\revtext{
The authors use a logistic function with a scale parameter $r$,
the carrier capacity parameter $K$ and the initial condition $c_0$ to model the
cumulative incidence $c(t)$: Assuming, (theoretically), the infection events
are observable. It is inaccurate to claim that r is the exponential growth
rate
}

We have added the following justification to the text:

\newtext{
We estimate the initial exponential growth rate $r$ of an epidemic from daily incidence by modeling the cumulative incidence $c(t)$ with a logistic function \citep{ma2014estimating}:
\begin{equation}
c(t) = \frac{K}{1 + \left[(K/c_0)-1\right]e^{-rt}}.
\end{equation}
While the exponential growth rate (i.e., the rate of change in log incidence) of the logistic function changes throughout an epidemic,
we focus strictly on estimating the \emph{initial} exponential growth rate (when $t \to -\infty$).
The method of estimating the initial exponential growth rate by fitting a logistic curve has been previously validated against simulations of stochastic compartmental models \citep{ma2014estimating}.
}

@@

Thank you for pointing this out. We made typographical errors in figure captions. We have corrected this.

\revtext{Line 145: "acquantanceships" $\rightarrow$ "acquaintanceships"}

Done.

\revtext{Figure 3 caption, last sentence: has recovered $\rightarrow$ have recovered}

### 2020-05-31 doc/network.tex
ready to submit

While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observaton.
The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infection that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproduction number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

@@

\jd{How do you feel about rewriting this paragraph? It sounds like we introduce space to understand effects of multiple contacts; in fact, the opposite is true; we got into the effects of multiple contacts because we were interested in space.}
Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals from an individual to have a smaller mean than their mean intrinsic generation interval.
To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquaintanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for simplicity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating the individual-level kernel over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma \tau} - e^{-\sigma \tau}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with constant transmission rate during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$). 
Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using the egocentric distribution to link $r$ and $\Rini$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
we briefly describe this relationship in Section~\ref{egosection}.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the contact results in infection.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution;
since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
Since spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles), we rely on simulations to validate this prediction.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describe the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describe the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

Here, we investigate two methods for correcting for temporal bias in aggregated generation-interval data (see Methods for details).
We refer to the first method as the population-level method as it relies on realized generation intervals aggregated across the entire population.
When an epidemic is growing exponentially, right-truncation causes the aggregated generation interval to be discounted by the exponential growth rate (\eref{exp});
hence, we can ``undo'' the truncation by exponentially weighting the aggregated generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.

We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector (\eref{nonhomogeneous}); 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the truncation process in the realized generation intervals.
For both methods, mean and coefficient of variation (CV) of generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of initial reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 75 infected individuals,
as well as the estimates of the initial reproductive number calculated from the initial forward generation intervals, which we define as the generation intervals for all infections caused by the first 75 infected individuals (\fref{cmp}).
For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting.
We do not expect under-reporting to affect the inference of generation-interval distributions (Appendix A.3) unless there are systematic biases in the observation process.
On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the reproductive number \cite{britton2019estimation}.
We do not pursue these directions in this study.

As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the initial forward generation-interval distribution on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In the Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

To simulate epidemics on a realistic network, we use the `condensed matter physics' network from the Stanford Large Network Dataset Collection \citep{leskovec2016snap}.
This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
It consists of 23133 nodes and 93497 edges.
The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the reproductive number.

>>>>

While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observation.
The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproduction number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

@@

The effects of spatial structure on realized generation intervals can be understood in terms of effect of multiple contacts.
Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals from an individual in a spatially structured population to have a smaller mean than their mean intrinsic generation interval.
To explore the effects of spatial structure on realized generation intervals, we relax our assumption that the population is homogeneous.
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquaintanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for simplicity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating the individual-level kernel over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of the basic reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

For example, consider a susceptible-exposed-infected-recovered (SEIR) model, which assumes that latent and infectious periods are exponentially distributed.
The intrinsic generation-interval distribution that corresponds to this model can be written as \citep{yan2008separate, Champredon2018equivalence}:
\begin{equation}
g(\tau) = \frac{\sigma \gamma}{\sigma - \gamma} \left(e^{-\gamma \tau} - e^{-\sigma \tau}\right),
\end{equation}
where $1/\sigma$ and $1/\gamma$ are mean latent and infectious periods, respectively.
Assuming a constant per-pair contact rate of $\lambda$ for any pair, we obtain the following egocentric generation-interval distribution:
\begin{equation}
\hat{g}(\tau) = \frac{\sigma (\gamma + \lambda)}{\sigma - (\gamma + \lambda)} \left(e^{-(\gamma + \lambda)\tau} - e^{-\sigma \tau}\right)
\end{equation}
In this case, with constant transmission rate during the infection period, the effect of accounting for pairwise contacts is the same as an increase in the recovery rate (by the amount of the per-pair contact rate $\lambda$). 
Infecting a susceptible contact is analogous to recovery because the contactee cannot be infected again --- the infector cannot no longer transmit infection even if they are infectious (effectively losing infectiousness).
Therefore, the resulting egocentric generation-interval distribution is equivalent to the intrinsic generation-interval distribution with mean latent period of $1/\sigma$ and mean infectious period of $1/(\gamma + \lambda)$.
In practice, directly using the egocentric distribution to link $r$ and $\Rini$ using the Euler-Lotka equation is unrealistic because it requires that we know the per-pair contact rate. 
Instead, the per-pair contact rate can be inferred from the growth rate $r$, assuming that mean and variance of the degree distribution of a network is known (see \cite{trapman2016inferring} supplementary material section 1.4.2);
we briefly describe this relationship in Section~\ref{egosection}.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all infectious contacts by a focal individual, regardless of whether the contact results in infection.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions have a shorter mean than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

In a large homogeneously mixing population, the initial forward generation-interval distribution is equivalent to the intrinsic distribution and provides the correct link between the exponential growth rate $r$ and the initial reproduction number \Rini (see \eref{forward}).
In a nonhomogeneous population, the initial forward generation-interval distributions are subject to spatial effects and, therefore, are different from the intrinsic distribution.
Since spatial effects have the same effect on how the epidemic spreads as they do on realized generation intervals, we expect the initial forward generation-interval distributions, which implicitly account for the spatial structure, to correctly link $r$ and $\Rini$ through the Euler-Lotka equation (\eref{euler}).
Spatial effects on realized generation intervals are generally expected to be analytically intractable, even in simple networks (e.g., see \cite{tomba2010some} for discussion regarding the realized generation intervals in a household with one infector and two susceptibles); therefore, we rely on simulations to validate this prediction.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

Here, we investigate two methods for correcting for temporal bias in aggregated generation-interval data (see Methods for details).
We refer to the first method as the population-level method as it relies on realized generation intervals aggregated across the entire population.
When an epidemic is growing exponentially, right-truncation causes the aggregated generation interval to be discounted by the exponential growth rate (\eref{exp});
hence, we can ``undo'' the truncation by exponentially weighting the aggregated generation-interval distribution \citep{nishiura2010time, tomba2010some, britton2019estimation}:
\begin{equation}
f_0(\tau) \propto a_0(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.

We refer to the second method as the individual-level method because it relies on individual contact information.
We model each infection as a non-homogeneous Poisson process arising from the infector (\eref{nonhomogeneous}); 
incorporating information about time of infection of an infector, time of infection of an infectee, and time since the beginning of an epidemic allows us to explicitly model the truncation process in the realized generation intervals.
For both methods, mean and coefficient of variation (CV) of the initial forward generation-interval distributions are estimated by maximum likelihood; the inferred generation-interval distributions are then used to estimate the initial reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of initial reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 75 infected individuals,
as well as the estimates of the initial reproductive number calculated from the empirical initial forward generation intervals, which we define as the generation intervals for all infections caused by the first 75 infected individuals (\fref{cmp}).
For simplicity, we assume that realized generation intervals are observed without error and assume that there is no under-reporting of generation intervals.
We do not expect under-reporting to affect the inference of generation-interval distributions (Appendix A.3) unless there are systematic biases in the observation process.
On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estimates of the initial reproductive number \cite{britton2019estimation}.
We do not pursue these directions in this study.

As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the aggregated generation intervals from contact tracing (via Euler-Lotka) severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the empirical initial forward generation-interval distribution on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In the Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

To simulate epidemics on a realistic network, we use the `condensed matter physics' network from the Stanford Large Network Dataset Collection \citep{leskovec2016snap}.
This graph describes co-authorship among anyone who submitted a paper to Condensed Matter category in the arXiv between January 1993 and April 2003 \citep{leskovec2007graph}.
It consists of 23133 nodes and 93497 edges.
The same network was used by \cite{trapman2016inferring} to study how network structure affects the estimate of the basic reproductive number.
